---
title: "demographic_assumptions_model"
format: html
editor: visual
---

## Demographic assumptions model

Home delivery (default):

-   older people (age 65+)

-   higher-income households

-   possibly suburban/less mobile populations

Service point pickup:

-   Middle-aged adults (25-64)

-   people with moderate income

-   likely to combine pickup with errrands

Automated Parcel Lockers (APLs)

-   younger adults (15-44)

-   tech-savy and busy people

-   low-to-middle income and urban areas

-   areas with more independence

Install and load packages:

```{r}
# install.packages(c("dplyr", "readxl"))
library(dplyr)
library(readxl)
library(ggplot2)
```

Load the cleaned CBS Squares data and prepare dataset:

```{r}
# Load the Excel file
data <- read_excel("/Users/app/Desktop/case-2-main/CBS_Squares_cleaned.xlsx", skip = 1)

# Rename columns for clarity
data <- data %>%
  rename(
    square = 1,
    population = Population,
    age_0_14 = `Age0.14`,
    age_15_24 = `Age15.24`,
    age_25_44 = `Age25.44`,
    age_45_64 = `Age45.64`,
    age_65_plus = `Age65.`,
    income_bracket = `Median.household.income`,
    x = X,
    y = Y
  )

# Convert numeric columns
cols_to_numeric <- c("population", "age_0_14", "age_15_24", "age_25_44",
                     "age_45_64", "age_65_plus", "x", "y")
data[cols_to_numeric] <- lapply(data[cols_to_numeric], as.numeric)
```

Define the city center of Maastricht (centroid of points 4387, 5499, 4431, 4987):

```{r}
city_center <- c(x = mean(c(55134, 51393, 66641, 67529)),
                 y = mean(c(39885, 53765, 48763, 44080)))
```

Closer to center = urban, further from center = suburban

Compute the distance to city center (urban vs suburban proxy):

```{r}
data <- data %>%
  mutate(
    dist_to_center = sqrt((x - city_center["x"])^2 + (y - city_center["y"])^2),
    urban = ifelse(dist_to_center < 25000, "urban", "suburban")
  )
```

Assign delivery preference based on assumptions:

```{r}
data <- data %>%
  rowwise() %>%
  mutate(
    delivery_preference = case_when(
      # APL preference
      (age_15_24 + age_25_44) > population * 0.3 &
        income_bracket %in% c("20-40 below middle", "20-60 below middle to middle", "60-80 above middle") &
        urban == "urban" ~ "apl",

      # Service point preference
      age_45_64 > population * 0.25 &
        income_bracket %in% c("60-80 above middle", "60-100 above middle to high") ~ "service_point",

      # Home delivery preference
      age_65_plus > population * 0.2 |
        income_bracket %in% c("80-100 high", "60-100 high") |
        urban == "suburban" ~ "home",

      # Default
      TRUE ~ "home"
    )
  ) %>%
  ungroup()

# Summary of preferences
table(data$delivery_preference)

# Save results
 write.csv(data, "/Users/app/Desktop/delivery_preferences_by_square.csv", row.names = FALSE)
```

```{r}
summary(data$dist_to_center[data$delivery_preference == "apl"])

```

## Visualisations 

```{r}
ggplot(data, aes(x = x, y = y, color = delivery_preference)) +
  geom_point(size = 2, alpha = 0.8) +
  scale_color_manual(values = c("home" = "firebrick", "service_point" = "steelblue", "apl" = "forestgreen")) +
  theme_minimal() +
  labs(
    title = "Predicted Delivery Preference by CBS Square (Maastricht)",
    x = "X Coordinate", y = "Y Coordinate", color = "Delivery Type"
  )
```
